Gyrochronal and Lithium Ages for Stars and Planets Observed by Kepler

### 2023-06-03 paper/ms.tex
update ms

\documentclass[12pt,twocolumn,tighten]{aastex63}
% upon AAS submission
%\documentclass[12pt,twocolumn,tighten,linenumbers]{aastex63}
%\documentclass[12pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
% drafting / arxiv
%\turnoffedit

\usepackage{apjfonts}
\usepackage{url}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{amsmath,amstext,amssymb}
\usepackage[caption=false]{subfig} % for subfloat
\usepackage{bm} % for bold hat accents
\usepackage{xcolor, fontawesome}
\usepackage{color}

\newcommand{\rprs}{{$R_p/R_{\star}$}}
\newcommand{\teff}{T$_{\textrm{eff}}$}
\newcommand{\gcc}{{g\,cm$^{-3}$}}
\newcommand{\rstar}{{$R_\star$}}
\newcommand{\mearth}{{M$_\oplus$}}
\newcommand{\rearth}{{R$_\oplus$}}
\newcommand{\rsun}{{R$_\odot$}}
\newcommand{\msun}{{M$_\odot$}}

@@

\newcommand{\berkeley}{Astronomy Department, University of California, Berkeley, CA 94720, USA}
\newcommand{\caltech}{Department of Astronomy, MC 249-17, California Institute of Technology, Pasadena, CA 91125, USA}

@@

%%%%%%%%%
% STARS %
%%%%%%%%%
% number of stars monitored by Kepler (quoting Santos19/21)
\newcommand{\nkeplerstars}{$\approx$160{,}000}
% fraction of kepler stars w/ rotation periods
\newcommand{\nstarswithprot}{{\bf $\sim$50{,}000}}
% fraction of stars with rotation periods, with teff/logg in Berger+20
\newcommand{\fracstarswithprotwithbtwenty}{{\bf $\sim$94\%}}
% complement of above
\newcommand{\fracstarswithprotwithoutbtwenty}{{\bf $\sim$6\%}}

%%%%%%%%%%%
% PLANETS %
%%%%%%%%%%%
% number of KOIs in Q1-Q17 DR25 table
\newcommand{\nkois}{{\bf $\sim$8{,}000}} %FIXME precise
% number of KOIs in Q1-Q17 DR25 table that are not known false positives
\newcommand{\nkoisnofp}{{\bf $\sim$4{,}000}} %FIXME precise
% number of KOIs in Q1-Q17 DR25 table that have rotation periods
\newcommand{\nkoiswithprot}{{\bf $\sim$2{,}000}}
% number of KOIs in Q1-Q17 DR25 table that are not known false positives that have rotation periods
\newcommand{\nkoisnofpwithprot}{{\bf $\sim$1{,}000}}

\newcommand{\fracplanetswithprotwithbtwenty}{{\bf $\sim$94\%}}

\newcommand{\nstarwithages}{NNN}

\defcitealias{McQuillan_2014}{M14}
\defcitealias{Mazeh_2015}{M15}
\defcitealias{Santos_2019}{S19}
\defcitealias{Santos_2021}{S21}

@@

\title{gyro-johannes}

@@

\input{authors.tex}

\begin{abstract}
  We report rotation- and lithium-based ages for stars observed by
  the primary Kepler mission.  Stellar rotation periods are collected from
  previous work; lithium is measured using both new and archival HIRES
  spectra.
  For the \nstarwithages\ FGK stars with reported ages, the relative statistical
  precisions $\sigma_t / t$ are XX +YY -ZZ\%, which is a factor of X.X
  +Y.Y -Z.Z more precise than state-of-the-art isochronal ages for the
  same stars.  The results include XXX planets younger than 1\,Gyr,
  and YYY planets younger than 0.5\,Gyr.
  In the base sample of dwarf FGK stars observed by
  Kepler, we find that the intrinsic age distribution is(/not)
  consistent with a uniform distribution between X and Y billion years
  (Gyr).  In the planet sample, we find a consistent(/inconsistent)
  distribution at $N$-sigma.  Analyzing the completeness of the Kepler
  detection pipeline, we interpret the majority of the change to be
  caused by selection effects (or astrophysics).  We also comment on a
  few previously reported evolutionary trends in the size evolution of
  close-in mini-Neptunes (2-4\,\rearth) and super-Earths
  (1-2\,\rearth), a few of which we recover at comparable
  ($\sim$3$\sigma$) significance to previous work, and one of which
  we note for the first time.
\end{abstract}

@@

\subsection{Stars}
\label{subsec:starsel}

@@

\subsection{Planets}
\label{subsec:planetsel}

@@

%FIXME: add a sentence or two describing the numbers after
%crossmatching against S19/S21 --

Crossmatching the Q1-Q17 DR25 KOI table against our stars with
rotation periods yields \nkoiswithprot\ KOIs with rotation periods, of which \nkoisnofpwithprot\ 
are not known false postives.

\subsection{Stellar Properties}

@@

As our default source for stellar temperatures and surface gravities, we adopt
the values from the Gaia-Kepler Stellar Properties Catalog (GKSP;
\citealt{Berger_2020a_catalog}).
The GKSP parameters are available for stars with ``AAA'' 2MASS photometry,
measured parallaxes in Gaia DR2,  and g-band photometry from either the KIC or
the Kepler-INT survey.
The parameters themselves were derived using \texttt{isoclassify} (CITE
Huber2017) to interpolate over the MIST isochrone grids (CITE Choi2016,
Dotter2016), given the SDSS $g$ and 2MASS $K_{\rm s}$ photometry, the Gaia DR2
parallexes, and $V$-band extinction from the (CITE) Green2019 reddening map.
The resulting stellar parameters are available for
\fracstarswithprotwithbtwenty\ of the \nstarswithprot\ Kepler stars with
rotation periods.
For the remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
temperatures and surface gravities, we adopt the values 
from the Gaia DR3 General Stellar Parametrizer from Photometry (GSP-Phot; CITE).
GSP-Phot uses the low-resolution Gaia BP and RP spectra, mean $G$ magnitudes,
and parallaxes, in concert with the PARSEC isochrones (CITE Tang2014, Chen2015,
Pastorelli2020) to infer the stellar effective temperatures, surface gravities,
metallicities, and line-of-sight extinctions.
TODO: plot uncertainties in teff/logg for both parameter sets.

% TODO: add that plot showing differences btwn authors

\subsection{Lithium}

%FIXME: left off here for discussion of HIRES spectra.

@@

@@

>>>>

% upon AAS submission
%\documentclass[12pt,twocolumn,tighten,linenumbers]{aastex63}
%\documentclass[12pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
% drafting / arxiv
\documentclass[11pt,twocolumn,tighten]{aastex63}
\turnoffedit

\usepackage{apjfonts}
\usepackage{url}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{amsmath,amstext,amssymb}
\usepackage{xcolor, fontawesome}
\usepackage{color}

\newcommand{\rprs}{{$R_p/R_{\star}$}}
\newcommand{\vsini}{{$V \sin i$}}
\newcommand{\teff}{T$_{\textrm{eff}}$}
\newcommand{\kms}{{km\,s$^{-1}$}}
\newcommand{\gcc}{{g\,cm$^{-3}$}}
\newcommand{\rstar}{{$R_\star$}}
\newcommand{\rhostar}{{$\rho_\star$}}
\newcommand{\mearth}{{M$_\oplus$}}
\newcommand{\rearth}{{R$_\oplus$}}
\newcommand{\rsun}{{R$_\odot$}}
\newcommand{\msun}{{M$_\odot$}}

@@

\newcommand{\eg}{e.g.,}

\newcommand{\mstype}{letter}

%%%%%%%%%%%%%%%%
% INSTITUTIONS %
%%%%%%%%%%%%%%%%
\newcommand{\caltech}{Department of Astronomy, MC 249-17, California Institute of Technology, Pasadena, CA 91125, USA}
\newcommand{\mitkavli}{MIT Kavli Institute and Department of Physics, 77 Massachusetts Avenue, Cambridge, MA 02139}
\newcommand{\berkeley}{Astronomy Department, University of California, Berkeley, CA 94720, USA}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%
% STARS %
%%%%%%%%%
% number of stars monitored by Kepler (quoting Santos19/21)
\newcommand{\nkeplerstars}{$\approx$160{,}000}
% fraction of kepler stars w/ rotation periods
\newcommand{\nstarswithprot}{{$\sim$50{,}000}}
% fraction of stars with rotation periods, with teff/logg in Berger+20
\newcommand{\fracstarswithprotwithbtwenty}{{$\sim$94\%}}
% complement of above
\newcommand{\fracstarswithprotwithoutbtwenty}{{$\sim$6\%}}

%%%%%%%%%%%
% PLANETS %
%%%%%%%%%%%
% number of KOIs in Q1-Q17 DR25 table
\newcommand{\nkois}{{$\sim$8{,}000}} %FIXME precise
% number of KOIs in Q1-Q17 DR25 table that are not known false positives
\newcommand{\nkoisnofp}{{$\sim$4{,}000}} %FIXME precise
% number of KOIs in Q1-Q17 DR25 table that have rotation periods
\newcommand{\nkoiswithprot}{{$\sim$2{,}000}}
% number of KOIs in Q1-Q17 DR25 table that are not known false positives that have rotation periods
\newcommand{\nkoisnofpwithprot}{{$\sim$1{,}000}}
% fraction of KOIs that are not FPs with B+20 parameters
\newcommand{\frackoisnofpwithprotwithbtwenty}{{$\sim$92\%}}

%%%%%%%%%%%
% LITHIUM %
%%%%%%%%%%%
% number of KOIs that are not known FPs that have a HIRES spectrum
\newcommand{\nkoiswithhires}{{1{,}500}}
% number of KOIs for which this work and Berger2018 overlap
\newcommand{\nbergeroverlap}{{500}}

%%%%%%%%
% AGES %
%%%%%%%%
\newcommand{\nkepstarswithages}{NNN}
\newcommand{\nkoisnofpwithages}{NNN}
\newcommand{\nkoisgoldwithages}{NNN}

%%%%%%%%%%%%%
% OFT-CITED %
%%%%%%%%%%%%%
\defcitealias{McQuillan_2014}{M14}
\defcitealias{Mazeh_2015}{M15}
\defcitealias{Santos_2019}{S19}
\defcitealias{Santos_2021}{S21}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\title{Gyrochronal and Lithium Ages for Stars and Planets Observed by Kepler}

@@

\author[0000-0002-0514-5538]{Luke~G.~Bouma}
\altaffiliation{51 Pegasi b Fellow}
\affiliation{\caltech}

\author{Authors to include (ORDER AFTER LGB NOT YET ASSIGNED MEANING)}

\author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
\affiliation{\caltech}

\author{Lynne~A.~Hillenbrand}
\affiliation{\caltech}

%
% HIRES Collaborators
%
\author[0000-0002-0531-1073]{H. Isaacson}
\affiliation{\berkeley}
%
\author[0000-0001-8638-0320]{A. W. Howard}
\affiliation{\caltech}
%
\author{HIRES observers to be named}
\affiliation{Various}

\begin{abstract}
  We report rotation- and lithium-based ages for stars and planets observed by
  the primary Kepler mission.  Stellar rotation periods are collected from
  previous analyses of the Kepler light curves; lithium is assessed directly from HIRES
  spectra.
  wWe report ages for \nkoisgoldwithages\ FGK stars with transiting planets, with relative statistical
  precisions $\sigma_t / t$ of XX +YY -ZZ\%; a factor of X.X
  +Y.Y -Z.Z more precise than state-of-the-art isochronal ages for the
  same stars.  The results include XXX planets younger than 1\,Gyr,
  and YYY planets younger than 0.5\,Gyr.
  In the base sample of dwarf FGK stars observed by
  Kepler, we find that the intrinsic age distribution is(/not)
  consistent with a uniform distribution between X and Y billion years
  (Gyr).  In the planet sample, we find a consistent(/inconsistent)
  distribution at $N$-sigma.  Analyzing the completeness of the Kepler
  detection pipeline, we interpret the majority of the change to be
  caused by selection effects (or astrophysics).  We also comment on a
  few previously reported evolutionary trends in the size evolution of
  close-in mini-Neptunes (2-4\,\rearth) and super-Earths
  (1-2\,\rearth), a few of which we recover at comparable
  ($\sim$3$\sigma$) significance to previous work, and one of which
  we note for the first time.
\end{abstract}

@@

\subsection{Kepler stars with rotation periods}
\label{subsec:starsel}

@@

\subsection{Kepler objects of interest}
\label{subsec:planetsel}

@@

Crossmatching the Q1-Q17 DR25 KOI table against our stars with
rotation periods yields \nkoiswithprot\ KOIs with rotation periods, of which \nkoisnofpwithprot\ 
are not known false postives.
These objects will comprise the core data for our gyrochronology analysis.

\subsection{Lithium sample}
A final subcomponent of our analysis involves an age assessment based on
the equivalent width of the \ion{Li}{1} 6708\,\AA\ doublet.  For this
component of the work, we primarily rely on spectra acquired with the
High Resolution Echelle Spectrometer (HIRES;
\citealt{vogt_hires_1994}) on the Keck I 10m telescope.
The majority of these spectra are archival, and were acquired
as part of the California Keck Survey (CKS).
A subset were acquired by our team in our own observing programs.
%TODO quantify.
%TODO compare to berger2018?

These HIRES spectra are available for \nkoiswithhires\ of the
\nkoisnofp\ non-false-positive KOIs.
In the broader stellar sample, they are available for a much smaller 
fraction of the stars observed by Kepler, because they were generally only
acquired due to the presence of a transiting planet candidate.
Given this selection function,
we leverage these spectra in our analysis as a way to scrutinize the
ages of just the planet sample.

\section{Stellar Properties}

@@

As our default source for stellar temperatures and surface gravities, we adopt
the values from the Gaia-Kepler Stellar Properties Catalog (GKSP;
\citealt{Berger_2020a_catalog}).
The GKSP parameters are available for stars with ``AAA'' 2MASS photometry,
measured parallaxes in Gaia DR2,  and g-band photometry from either the KIC or
the Kepler-INT survey.
The parameters themselves were derived using \texttt{isoclassify} (CITE
Huber2017) to interpolate over the MIST isochrone grids (CITE Choi2016,
Dotter2016), given the SDSS $g$ and 2MASS $K_{\rm s}$ photometry, the Gaia DR2
parallexes, and $V$-band extinction from the (CITE) Green2019 reddening map.
The resulting stellar parameters are available for
\fracstarswithprotwithbtwenty\ of the \nstarswithprot\ Kepler stars
with rotation periods.
For the remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
temperatures and surface gravities, we adopt the values 
from the Gaia DR3 General Stellar Parametrizer from Photometry (GSP-Phot; CITE).
GSP-Phot uses the low-resolution Gaia BP and RP spectra, mean $G$ magnitudes,
and parallaxes, in concert with the PARSEC isochrones (CITE Tang2014, Chen2015,
Pastorelli2020) to infer the stellar effective temperatures, surface gravities,
metallicities, and line-of-sight extinctions.
In the planet sample,
\frackoisnofpwithprotwithbtwenty\ of the
the non-false-positive KOIs with rotation periods have parameters from
\citet{Berger_2020a_catalog}.

As a sanity check on our effective temperature scale, we then used the 
the CITE Green2019 reddening map, combined with the Gaia DR2 $G_{\rm
BP}-G_{\rm RP}$ color, to calculate effective temperatures using the
relation suggested by \citet{Curtis_2020}.
This latter effective temperature scale was demonstrated to have a 
scatter of $\approx$50\,K for stars near the main-sequence.
We find {\bf X, Y, Z.}
We ultimately favor the \citet{Berger_2020a_catalog} scale
because it uses 2MASS $K_{\rm s}$ photometry, and is tied to an actual
isochronal fit.

%TODO: plot uncertainties in teff/logg for both parameter sets.

% TODO: add that plot showing differences btwn the two teff scales?

@@

\subsection{Gyrochrone ages}
To calculate gyrochrone ages, we use \texttt{gyro-interp}
\citep{Bouma_2023}.  This framework was built to statistically address
the issue that, especially over the first gigayear of stellar
spin-down, stars with the same mass and same rotation period can have
very different ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The
gyrochrone age posterior should therefore incorporate this intrinsic
population-level scatter into the precision with which the age can be
measured.  The three main assumptions behind the \texttt{gyro-interp}
framework are as follows.  {\it 1)} The stars are on the main
sequence, {\it 2)} they do not have any binary companions,
and {\it 3)} the rotation period distribution
for field stars of a given mass and age is identical to that of
the calibration open clusters (metallicity differences, for instance, are
ignored).  
Regarding the former two assumptions:
pre-main-sequence and post-main-sequence evolution introduce
variations in the stellar angular momentum budget.  Similarly, tidal
effects might systematically spin up (or down) stars in binaries.

Using our adopted effective temperatures and rotation periods, we ran 
\texttt{gyro-interp} over an age grid of 0 to 5\,Gyr, linearly spaced
over 500 grid points.
\texttt{gyro-interp} itself is only calibrated out to 4\,Gyr, the
adopted age of the M67 cluster \citep[see][]{Dungee_2022,Gruner_2023}.
We used the spin-down law between the reference open clusters by
interpolating using piecewise cubic hermite interpolating polynomials,
and integrated Equation~1 of \citep{Bouma_2023} at the default grid
resolution.
This yielded age posterior probabilities over our requested age grid
for \nkepstarswithages\ Kepler stars, and \nkoisnofpwithages\
non-false-positive KOIs.

\subsection{Lithium ages}
To measure the lithium equivalent widths for the \nkoiswithhires\ KOIs
with HIRES spectra, we use the procedure described by
\citet{Bouma_2021}.
The stars of interest for this work are FGK stars, and so the
continuum in the vicinity of the
\ion{Li}{1} 6708\,\AA\ feature is well-defined.
We therefore numerically integrate a single best-fit Gaussian over a
local window, and estimate the uncertainties on the line width through
a Monte Carlo procedure that bootstraps against the local scatter in the
continuum.
Generally speaking, this approach does not correct for the neighboring
\ion{Fe}{1} 6707.44\,\AA\ blend.

To evaluate the accuracy and precision of this approach, after
applying an initial iteration of this method on the \nkoiswithhires\
KOIs with HIRES spectra, we compared our lithium equivalent widths
with those reported by \citet{Berger_2018}.
For the \nbergeroverlap\ stars in both samples, 
we found {\bf X, Y, Z.}

To calculate lithium ages, we then use EAGLES \citep{Jeffries_2023}.
{\bf TODO: explain why}
%TODO: explain why.

\subsection{Quality flags}
In our core gyrochrone analysis, we calculate gyrochrone age
posteriors for all the stars with reported rotation periods and
effective temperatures.  To select the stars for which we expect the
gyrochronal age to be valid, we then build a set of quality flags, and
apply them in the sub-analyses that follow.  We defined these filters
as follows, each of which is included in Tables~\ref{tab:stars}
and~\ref{tab:planets}.

{\it Subgiants}---We flagged stars as possible subgiants if they had
$\log g < 4.2$.  We determined this threshold by examining the
relevant flag from \citet{berger_2018_radii_evolnstates},
cross-matched against \citet{Berger_2020a_catalog}.
\citet{berger_2018_radii_evolnstates} performed this classification
using the PARSEC evolutionary models, stellar radii, and DR25 stellar
properties catalog effective temperatures \citep{Mathur_2017}.

{\it Photometric binaries}---We plotted color--absolute magnitude
diagrams using the Gaia DR2 photometry, in $M_{\rm G}$ vs. $(G_{\rm
BP} - G_{\rm RP})_0$, and $(G - G_{\rm RP})_0$.  We then manually drew
loci to flag over or under-luminous stars in each diagram, typically
more than $\approx$0.5 magnitudes above or below the main sequence.
%TODO refer to a figure!!!

{\it RUWE}---Following recommendations in the Gaia DR3
documentation\footnote{\url{https://gea.esac.esa.int/archive/documentation/GDR2/Gaia_archive/chap_datamodel/sec_dm_main_tables/ssec_dm_ruwe.html}},
we flagged stars with a Gaia DR3 renormalized unit weight error (RUWE)
exceeding 1.4 as possible binaries.  Such astrometric outliers can be
either bonafide astrometric binaries, or more often are marginally
resolved point sources for which the single-source PSF model used in
the Gaia pipeline induces spurious astrometric scatter.

{\it Crowding}---We searched the Gaia DR3 point source catalog for
stars within 1 Kepler pixel (4$''$) of every target star.
While many such companions are not physically associated with the
target star, their presence can bias stellar parameters and stellar
rotation period measurements.
We therefore flag any stars with neighbors down to 10$|times$ the 
brightness of the target star within this region ($\Delta G < 2.5$).

{\it Gaia DR3 Non-Single-Stars}---The Gaia DR3
\texttt{non\_single\_star} column in the \texttt{gaia\_source} table
flags known eclipsing, astrometric, and spectroscopic binaries.  We
directly include this column.

{\it Final calibration sample}---{\bf todo: mention any extra flags
from Berger2020, or perhaps in the KOI catalog itself, or perhaps the
amplitude cuts you might want to apply to the rotation period catalog}
%The combination of the filters
%described above yields the set of stars that show no evidence for
%binarity or crowding.  However, some of the rotation period analyses
%in Table~\ref{tab:clusters} include additional relevant quality flags.
%For instance, light curves showing multiple photometric periods can
%indicate unresolved binarity.  We used all relevant filters available
%from the original authors if they were designed to select single stars
%with reliable rotation periods.  The final combination of these
%filters with our own flag for possible binarity yields our sample of
%benchmark rotators.

@@

\end{document}
